---
title: "SEM lab 4"
output:
  html_document: default
  html_notebook: default
---

### 1. RAM (the kind of model we are using), is implemented with Symmetric (S), Asymmetric (A), means (M) and a filter (F) matrix

2. Build our old friend m1: the model of mpg from ?umxRAM

```{r}
#install.packages("umx", dependencies = TRUE)
library(umx)
# 1. For convenience, list up the manifests you will be using
selVars = c("mpg", "wt", "disp")

# 2. Create an mxData object
myCov = mxData(cov(mtcars[,selVars]), type = "cov", numObs = nrow(mtcars) )

# 3. Create the model (see ?umxPath for more nifty options)
m1 = umxRAM("tim", data = myCov,
	umxPath(c("wt", "disp"), to = "mpg"),
	umxPath(cov = c("wt", "disp")),
	umxPath(var = c("wt", "disp", "mpg"))
)

```


### 3. Type: m1@matrices$S

What is this for?

```{r}
m1@matrices$S

```


### This is the object that stores Symmetric (2-headed) paths

What do you see (~5 things to note)?

* Is has a values matrix, which is where path values live

* It has a matrix storing whether a cell is free or fixed

* It has matrices for the lower (lbound) and upper (ubound) limits on values a cell can take
(may be NA)

* It has a matrix for labels each cell takes (maybe NA)

### Which cells (paths) in S are free for our model?

residuals (on the diagonal)


### 4. Type: m1@matrices$A
What do you see?
This is the matrix of Asymmetric paths

```{r}
m1@matrices$A
```


* Which cells (paths) in A are free for our model?
paths from the latent variable to each manifest


## 5. Look up the help for ?mxConstraint

```{r}
?mxConstraint
```


6. Run the example model: What is it doing?
It is forcing the matrix "K" to be equal to the matrix "limit"

```{r}
library(OpenMx)

#Create a constraint between MxMatrices 'A' and 'B'
constraint <- mxConstraint(A > B, name = 'AdominatesB')

# Constrain matrix 'K' to be equal to matrix 'limit'

model <- mxModel(model="con_test", 
    mxMatrix(type="Full", nrow=2, ncol=2, free=TRUE, name="K"),
    mxMatrix(type="Full", nrow=2, ncol=2, free=FALSE, name="limit", values=1:4),
    mxConstraint(K == limit, name = "Klimit_equality"), 
    mxAlgebra(min(K), name="minK"), 
    mxFitFunctionAlgebra("minK")
)


fit <- mxRun(model)
fit$matrices$K$values

#      [,1] [,2]
# [1,]    1    3
# [2,]    2    4

# Constrain both free parameters of a matrix to equality using labels (both are set to "eq")
equal <- mxMatrix("Full", 2, 1, free=TRUE, values=1, labels="eq", name="D")

# Constrain a matrix element in to be equal to the result of an algebra
start <- mxMatrix("Full", 1, 1, free=TRUE,  values=1, labels="param", name="F")
alg   <- mxAlgebra(log(start), name="logP")

# Force the fixed parameter in matrix G to be the result of the algebra
end   <- mxMatrix("Full", 1, 1, free=FALSE, values=1, labels="logP[1,1]", name="G")
```



### 7. We haven't covered it yet, but look up ?mxAlgebra

```{r}
?mxAlgebra
```


### Why might you need an algebra in a model?
* To compute new variables: like a %*% = A
* To compute the expected covariances of our model
* To compute the expected covariances of our model

### 8. Look up the help for ?mxCheckIdentification
```{r}
?mxCheckIdentification
```

### 9. Run the example model

```{r}
require(OpenMx)

data(demoOneFactor)
manifests <- names(demoOneFactor)
latents <- "G1"
model2 <- mxModel(model="One Factor", type="RAM",
      manifestVars = manifests,
      latentVars = latents,
      mxPath(from = latents[1], to=manifests[1:5]),
      mxPath(from = manifests, arrows = 2, lbound=1e-6),
      mxPath(from = latents, arrows = 2, free = FALSE, values = 1.0),
      mxData(cov(demoOneFactor), type = "cov", numObs=500)
)
fit2 <- mxRun(model2)

id2 <- mxCheckIdentification(fit2)
id2$status
# The model is locally identified

# Build a model from the solution of the previous one
#  but now the factor variance is also free
model2n <- mxModel(fit2, name="Non Identified Two Factor",
      mxPath(from=latents[1], arrows=2, free=TRUE, values=1)
)

mid2 <- mxCheckIdentification(model2n)
mid2$non_identified_parameters
# The factor loadings and factor variance
#  are not identified.
```


What was the effect of setting the latent variable's variance (model2n)?

* Model no longer identified
* What happens if you run the model
* Some models will still run, but can give different results each time, others can't be run

### 10. Use path tracing rules

What is the effect of a two-path connection .5 and .3?

.15 (the product of the connections in the pathway)

Draw a model a bit like the one for reading on the poster on the wall, but simpler :-)

put estimated values on it.

Use the path tracing rules to show the total influence of a latent variable in your model on a
particular manifest

https://en.wikipedia.org/wiki/Path_analysis_(statistics)#Path_tracing_rules

### 11. Run the example in ?umxMI (umx's modification indices function)

```{r}
?umxMI
require(umx)
data(demoOneFactor)
latents  = c("G")
manifests = names(demoOneFactor)[1:3]
df = mxData(cov(demoOneFactor[,manifests]), type = "cov", numObs = 500)
m4 <- umxRAM("One Factor", data = df,
	umxPath(latents, to = manifests),
	umxPath(var = manifests),
	umxPath(var = latents, fixedAt = 1)
)
umxMI(m4, full=FALSE)
```


The output shows paths that would appear to lead to a better fit.
In this case, none are very important, but try adding one of them to a copy of this model,
and compare the fit

note: This is COMPLETELY POST HOC

### 12. Read the warnings in ?mxMI

```{r}
?mxMI
```


Explain to your lab mates why modification indices risk unreplicable models

Can you think of a situtation where they might be of value?

How would you test a modified model?

### 13. Note: MI is a very limited, automated method which looks at evidence for paths given the
model. It can't invent new (potentially MUCH better) models and suggest those. That is what AI
tries to do.

# Extra credit

1. Run the 1-factor model (first model in examples in ?umx)

2. What does m1$F show you?
Hint "F" stands for filter
The latent variables appear only as columns, not rows
This allows the model to know what variables are latent

3. Build a twin model to practice for next week
?umxACE

# Refs
1. David Kenny page
http://davidakenny.net/cm/fit.htm
2. umx home page
http://tbates.github.io
3. OpenMx home page
http://openmx.ssri.psu.edu